A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2018-11-01 doc/large_strain.tex
do not use \nabla in algorithms (we don't use it in theoretical part)

We now introduce a FE triangulation $\mathcal{B}^h_0$ of $\mcl B_0$ and
the associated FE space of continuous piecewise elements with polynomial space of fixed degree. % : $V^h \subset H^1 (\mcl B_0)$.
The displacement fields are given in
a vector space spanned by standard vector--valued FE basis functions $\gz N_i(\gz x)$ (e.g. polynomials with local support on a patch of elements):
\begin{alignat}{2}
       \gz u^h &=:  \sum_{i \in \mcl I} u^i \gz N_i (\gz X) \quad \quad \quad
\delta \gz u^h &&=: \sum_{i \in \mcl I} \delta u^i \gz N_i (\gz X) \,,
\end{alignat}
where the superscript $h$ denotes that this representation is related to the FE mesh with size function $h(\gz X)$ and $\mcl I$ is the set of unknown degrees of freedom (DoF).
%{\color{red} For the sake of this study, $\gz N_i (\gz X)$ are such that only one component of the vector-valued shape function is non-zero for each $i$.}

@@

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{RHS FE vector $\gz x$, current FE solution $\overline{\gz u^h}$,
  cached $c_1 := \mu - 2 \lambda \log(J)$ for each cell and quadrature point}
  \Output{action of the FE tangent \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        evaluate the following tensors at each quadrature point using sum factorization \;
        $\gz \nabla_X \gz u^h$ \tcp*{2nd order}
        $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
        $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\gz F = \gz I + \gz \nabla_X \gz u^h$ \tcp*{2nd order}
          evaluate $J = \rm{det}(\gz F)$ \tcp*{scalar}
          evaluate $\gz b = \gz F \cdot \gz F^T$ \tcp*{2nd order symmetric}
          evaluate $\gz \tau = \mu \gz b - c_1 \gz I$ \tcp*{2nd order symmetric}
          evaluate $\boldsymbol{\mathcal{G}} \gz g := \gz g \cdot \gz \tau/J$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
          evaluate $\boldsymbol{\mathcal{C}}\gz g_s := \left[2 c_1 \gz g_s + 2 \lambda\, \rm{tr}( \gz g_s) \gz I \right]/J$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}} \gz g_s$ for contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
        }
      evaluate queued contractions using sum factorization and distribute results to the destination vector
  }
  \caption{Matrix-free tangent cell operator: scalar cache}
  \label{alg:mf_scalar}
\end{algorithm}

@@

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{RHS FE vector $\gz x$,
  cached
  $c_1 := 2\left[\mu - 2 \lambda \log(J)\right]/J$,
  $c_2 := 2\lambda/J$ and $\gz \tau/J$ for each cell and quadrature point,
  evaluated based on $\overline{\gz u^h}$}
  \Output{action of the FE tangent \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        evaluate the following tensors at each quadrature point using sum factorization \;
        $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
        $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\boldsymbol{\mathcal{G}}\gz g := \gz g \cdot \left[\gz \tau/J \right]$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$ \;
          evaluate $\boldsymbol{\mathcal{C}} \gz g_s := c_1 \gz g_s + c_2 \, \rm{tr}(\gz g_s) \gz I$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}} \gz g_s$ contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}} \gz g_s $ \;
        }
        evaluate queued contractions using sum factorization and distribute results to the destination vector
  }
  \caption{Matrix-free tangent cell operator, variant 1: cache second-order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate referential quantities like $\gz F$ at when executed.}
  \label{alg:mf_tensor2}
\end{algorithm}

In order to avoid recalculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point. In this case, we also avoid the need to evaluate gradients of displacement field with respect to the undeformed configuration.
This algorithm also utilizes the chosen constitutive relationship in that the operation of $J \boldsymbol{\mathcal{C}}$ remains directly expressed using \eqref{eq:simplified_action}.
Therefore, the action of the material fourth-order tangent $J \boldsymbol{\mathcal{C}}$ on the second-order symmetric tensor $\gz \nabla_{sym} \gz x$ is cheap to evaluate.
To that end, we cache two scalars that depend on the Jacobian $J$.

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{RHS FE vector $\gz x$,
  cached $\gz \tau/J$ and $\boldsymbol{\mathcal{C}}$ for each cell and quadrature point,
  evaluated based on $\overline{\gz u^h}$}
  \Output{action of the FE tangent \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        evaluate the following tensors at each quadrature point using sum factorization \;
        $\gz g_s := \gz \nabla_{sym} \gz x$ \tcp*{2nd order symmetric}
        $\gz g := \gz \nabla \gz x$ \tcp*{2nd order}
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\boldsymbol{\mathcal{G}} \gz g := \gz g \cdot \left[\gz \tau/J\right]$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\gz \nabla \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
          evaluate $\boldsymbol{\mathcal{C}}\gz g_s := \boldsymbol{\mathcal{C}}\gz : \gz g_s$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}}\gz g_s$ for contraction $\gz \nabla_{sym} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
        }
        evaluate queued contractions using sum factorization and distribute results to the destination vector
  }
  \caption{Matrix-free tangent cell operator: cache fourth-order material tangent $\boldsymbol{\mathcal{C}}$ and Kirchhoff stress $\gz \tau$.}
  \label{alg:mf_tensor4}
\end{algorithm}

Finally, Algorithm \ref{alg:mf_tensor4} represents the most general case which does not assume any form of the material tangent tensor.
In this case, we have to cache the fourth-order symmetric tensor $\boldsymbol{\mathcal{C}}$ and the Kirchhoff stress $\gz \tau$ for each element and quadrature point, and perform the double contraction with the second-order symmetric tensor $\gz \nabla_{sym} \gz x$ on-the-fly.

>>>>

We now introduce a FE triangulation $\mathcal{B}^h_0$ of $\mcl B_0$ and
the associated FE space of continuous piecewise elements with polynomial space of fixed degree. % : $V^h \subset H^1 (\mcl B_0)$.
The displacement fields are given in
a vector space spanned by standard vector--valued FE basis functions $\gz N_i(\gz x)$ (e.g. polynomials with local support on a patch of elements):
\begin{alignat}{2}
       \gz u^h &=:  \sum_{i \in \mcl I} u_i \gz N_i (\gz X) \quad \quad \quad
\delta \gz u^h &&=: \sum_{i \in \mcl I} \delta u_i \gz N_i (\gz X) \,,
\end{alignat}
where the superscript $h$ denotes that this representation is related to the FE mesh with size function $h(\gz X)$ and $\mcl I$ is the set of unknown degrees of freedom (DoF).
%{\color{red} For the sake of this study, $\gz N_i (\gz X)$ are such that only one component of the vector-valued shape function is non-zero for each $i$.}

@@

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{Source FE vector $\gz x$, current FE solution $\overline{\gz u^h}$,
  cached $c_1 := \mu - 2 \lambda \log(J)$ for each cell and quadrature point}
  \Output{action of the FE tangent \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        evaluate the following tensors at each quadrature point using sum factorization \;
        $\Grad \overline{\gz u^h}$ \tcp*{2nd order}
        $\gz g_s := \overline{\grad^s} \gz x$ \tcp*{2nd order symmetric}
        $\gz g := \overline{\grad} \gz x$ \tcp*{2nd order}
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\gz F = \gz I + \Grad \overline{\gz u^h}$ \tcp*{2nd order}
          evaluate $J = \rm{det}(\gz F)$ \tcp*{scalar}
          evaluate $\gz b = \gz F \cdot \gz F^T$ \tcp*{2nd order symmetric}
          evaluate $\gz \tau = \mu \gz b - c_1 \gz I$ \tcp*{2nd order symmetric}
          evaluate $\boldsymbol{\mathcal{G}} \gz g := \gz g \cdot \gz \tau/J$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
          evaluate $\boldsymbol{\mathcal{C}}\gz g_s := \left[2 c_1 \gz g_s + 2 \lambda\, \rm{tr}( \gz g_s) \gz I \right]/J$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}} \gz g_s$ for contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
        }
      evaluate queued contractions using sum factorization and distribute results to the destination vector
  }
  \caption{Matrix-free tangent cell operator: scalar cache}
  \label{alg:mf_scalar}
\end{algorithm}

@@

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{Source FE vector $\gz x$,
  cached
  $c_1 := 2\left[\mu - 2 \lambda \log(J)\right]/J$,
  $c_2 := 2\lambda/J$ and $\gz \tau/J$ for each cell and quadrature point,
  evaluated based on $\overline{\gz u^h}$}
  \Output{action of the FE tangent \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        evaluate the following tensors at each quadrature point using sum factorization \;
        $\gz g_s := \overline{\grad^s} \gz x$ \tcp*{2nd order symmetric}
        $\gz g := \overline{\grad} \gz x$ \tcp*{2nd order}
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\boldsymbol{\mathcal{G}}\gz g := \gz g \cdot \left[\gz \tau/J \right]$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$ \;
          evaluate $\boldsymbol{\mathcal{C}} \gz g_s := c_1 \gz g_s + c_2 \, \rm{tr}(\gz g_s) \gz I$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}} \gz g_s$ contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}} \gz g_s $ \;
        }
        evaluate queued contractions using sum factorization and distribute results to the destination vector
  }
  \caption{Matrix-free tangent cell operator, variant 1: cache second-order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate referential quantities like $\gz F$ at when executed.}
  \label{alg:mf_tensor2}
\end{algorithm}

In order to avoid recalculation of the Kirchhoff stress, Algorithm \ref{alg:mf_tensor2} caches its value for each element and quadrature point. In this case, we also avoid the need to evaluate gradients of displacement field with respect to the undeformed configuration.
This algorithm also utilizes the chosen constitutive relationship in that the operation of $J \boldsymbol{\mathcal{C}}$ remains directly expressed using \eqref{eq:simplified_action}.
Therefore, the action of the material fourth-order tangent $J \boldsymbol{\mathcal{C}}$ on the second-order symmetric tensor $\overline{\grad^s} \gz x$ is cheap to evaluate.
To that end, we cache two scalars that depend on the Jacobian $J$.

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{Source FE vector $\gz x$,
  cached $\gz \tau/J$ and $\boldsymbol{\mathcal{C}}$ for each cell and quadrature point,
  evaluated based on $\overline{\gz u^h}$}
  \Output{action of the FE tangent \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        evaluate the following tensors at each quadrature point using sum factorization \;
        $\gz g_s := \overline{\grad^s} \gz x$ \tcp*{2nd order symmetric}
        $\gz g := \overline{\grad} \gz x$ \tcp*{2nd order}
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\boldsymbol{\mathcal{G}} \gz g := \gz g \cdot \left[\gz \tau/J\right]$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
          evaluate $\boldsymbol{\mathcal{C}}\gz g_s := \boldsymbol{\mathcal{C}}\gz : \gz g_s$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}}\gz g_s$ for contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
        }
        evaluate queued contractions using sum factorization and distribute results to the destination vector
  }
  \caption{Matrix-free tangent cell operator: cache fourth-order material tangent $\boldsymbol{\mathcal{C}}$ and Kirchhoff stress $\gz \tau$.}
  \label{alg:mf_tensor4}
\end{algorithm}

Finally, Algorithm \ref{alg:mf_tensor4} represents the most general case which does not assume any form of the material tangent tensor.
In this case, we have to cache the fourth-order symmetric tensor $\boldsymbol{\mathcal{C}}$ and the Kirchhoff stress $\gz \tau$ for each element and quadrature point, and perform the double contraction with the second-order symmetric tensor $\overline{\grad^s} \gz x$ on-the-fly.
